---
title: "Custom time distributions (using pure R)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom time distributions (using pure R)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`{NEXTNetR}` provides a set of built-in time distributions that can be used as transmission and recovery times in simulations, see `help(time_distributions)`. Additional distributions can be defined by users through either R or C++ code. Here we discuss the simpler but much (order of magnitude!) slower R solution. See `vignette(custom_time_distributions_cpp)` for the faster but more complex C++ solution.

## Loading the NEXTNetR package

We start with loading the `{NEXTNetR}` package. If the package is not already installed, see the [website](https://oist.github.io/NEXTNetR/) for installation instructions. We also load the `{ggplot2}` and `{ggpubr}` packages for plotting and set a nice theme.

```{r setup}
library(NEXTNetR)
library(ggplot2)
library(ggpubr)
theme_set(theme_pubr())
```

## Implementing a custom time distribution in pure R

To implement a custom time distribution, we use `userdefined_time()`. This function allows us to provide arbitrary R functions that implement the functions listed in `help(time_functions)` for our distribution. Each function must take either a single argument `tau` or three arguments `tau`, `t` and `m`. At least two functions `survival` and `density` must be provided. by providing single-argument functions, we signal to *NEXTNet* that we only implement the base distribution and leave it to *NEXTNet* to derive the conditioned and modulated version of the distribution from that, see discussions in `help(time_functions)` and `help(userdefined_time)`.

We use `userdefined_time()` to create a function `mixture_time` which returns a mixture of the distributions listed in `times` with weights `weights`.

```{r}
mixture_time <- function(times, weights) {
  ud_time <- NULL
  ud_time <- userdefined_time(
    survival=function(tau)
      sum(sapply(times, time_survivalprobability, tau) * weights),
    density=function(tau)
      sum(sapply(times, time_density, tau) * weights),
    sample=function(tau) {
      i <- sample.int(n=length(weights), prob = weights, size=1)
      return(time_sample(1, times[[i]]))
    },
  )
}

```

We now use the newly created `mixture_time` function to create a mixture of an exponential distribution with mean 5 and a log-normal distribution with mean 200 and variance 500, weighted so that 30% of samples come from the exponential distribution.

```{r}
psi <- mixture_time(times=list(exponential_time(1/10),
                    lognormal_time(200, 500)), weights=c(0.3, 0.7))
```

To test this we sample from the newly created distribution using `time_sample`

```{r}
samples <- time_sample(1e3, psi, t=10, m=2)
```

and use `time_density` and `time_survivalprobability` to compare the distribution of our samples (in blue) to the theoretical distribution (in black) to check that everything works correctly

```{r warning=FALSE}
samples.ecdf <- ecdf(samples)
ggarrange(
  ggplot() +
    lims(x=c(0, 300)) +
    geom_histogram(data=data.frame(time=samples),
	               aes(x=time, y=after_stat(density)),
                   binwidth=2, fill="cornflowerblue") +
    geom_function(fun=time_density, n=1000,
                  args=list(timedistribution=psi, t=10, m=2)),
  ggplot() +
    lims(x=c(0, 300), y=c(0,1)) + labs(x='time', y='survival') +
    geom_function(fun=function(x) 1 - samples.ecdf(x),
	              color="cornflowerblue", linewidth=2) +
    geom_function(fun=time_survivalprobability, n=1000,
                  args=list(timedistribution=psi, t=10, m=2)),
  ncol=1
)
```

## Simulations using our custom distribution

Our custom distribution in simulations in the same way as built-in distributions, see `vignette(NEXTNetR)` for a step by step explanation of this

```{r}
nw <- erdos_renyi_network(1e3, 5)
sim <- simulation(nw, psi)
simulation_addinfections(sim, nodes=c(1), times=c(0.0))
events <- simulation_run(sim, stop=list(total_infected=300e3))
ggplot(events) +
  geom_line(aes(x=time, y=infected))
```
