---
title: "custom_time_distributions_r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{custom_time_distributions_r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`{NEXTNetR}` provides a set of built-in time distributions that can be used as transmission and recovery times in simulations, see `help(time_distributions)`. Additional distributions can be defined by users through either R or C++ code. Here we discuss the simpler but much (order of magnitude!) slower R solution. See `help(custom_time_distributions_r)` for the faster but more complex C++ solution.

## Loading the NEXTNetR package

We start with loading the `{NEXTNetR}` package. If the package is not already installed, see the [website](https://oist.github.io/NEXTNetR/) for installation instructions. We also load the `{ggplot2}` and `{ggpubr}` packages for plotting and set a nice theme.

```{r setup}
library(NEXTNetR)
library(ggplot2)
library(ggpubr)
theme_set(theme_pubr())
```

```{r}
mixture_time_r <- function(times, weights) {
  ud_time <- NULL
  ud_time <- userdefined_time(
    sample=function(tau) {
      i <- sample.int(n=length(weights), prob = weights, size=1)
      return(time_sample(1, times[[i]]))
    },
    sample_is_trinary=FALSE,
    survival=function(tau) {
      sum(sapply(times, time_survivalprobability, tau) * weights)
    },
    survival_is_trinary=FALSE,
    density=function(tau) {
      sum(sapply(times, time_density, tau) * weights)
    },
  )
}

```

```{r}
psi <- mixture_time_r(times=list(exponential_time(1/10), lognormal_time(200, 500)), weights=c(0.3, 0.7))
```

To test this we sample from the newly created distribution using `time_sample`

```{r}
samples <- time_sample(1e3, psi, t=10, m=2)
```

and use `time_density` and `time_survivalprobability` to compare the distribution of our samples (in blue) to the theoretical distribution (in black) to check that everything works correctly

```{r warning=FALSE}
samples.ecdf <- ecdf(samples)
ggarrange(
  ggplot() +
    lims(x=c(0, 300)) +
    geom_histogram(data=data.frame(time=samples), aes(x=time, y=after_stat(density)),
                   binwidth=2, fill="cornflowerblue") +
    geom_function(fun=time_density, n=1000,
                  args=list(timedistribution=psi, t=10,  m=2)),
  ggplot() +
    lims(x=c(0, 300), y=c(0,1)) + labs(x='time', y='survival') +
    geom_function(fun=function(x) 1 - samples.ecdf(x), color="cornflowerblue", linewidth=2) +
    geom_function(fun=time_survivalprobability, n=1000,
                  args=list(timedistribution=psi, t=10, m=2)),
  ncol=1
)
```


```{r}
nw <- erdos_renyi_network(1e3, 5)
sim <- simulation(nw, psi)
simulation_addinfections(sim, nodes=c(1), times=c(0.0))
events <- simulation_run(sim, stop=list(total_infected=300e3))
ggplot(events) +
  geom_line(aes(x=time, y=infected))
```