---
title: "Spatial and Temporal Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial and Temporal Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading the NEXTNetR package

We start with loading the `{NEXTNetR}` package. If the package is not already installed, see the [website](https://oist.github.io/NEXTNetR/) for installation instructions. We also load the `{ggplot2}` and `{ggpubr}` packages for plotting and set a nice theme.

```{r setup}
library(NEXTNetR)
library(ggplot2)
library(ggpubr)
theme_set(theme_pubr())
```

## Spatial Networks

Next we create a spatially embedded network representing a cubic lattice of size $50\times 50$, i.e. a network of nodes positioned at integral coordinates $(i,j)$ connected to their left, right, upper and lower neighbour. 

```{r}
L <- 50
nw <- cubiclattice2d_network(L);
```

For spatially embedded networks such as lattices, we can query the positions of nodes using `network_coordinates()`.`network_coordinates()` returns a $N \times d$ matrix, where $N$ is the number of nodes and $d$ the dimension of the embedding space, in our case 2. For convenience, we assign names to the coordinates, here 'x' and 'y'.

```{r}
N <- network_size(nw)
coords <- network_coordinates(nw, 1:N)
colnames(coords) <- c('x', 'y')
```

The coordinate matrix looks like this

```{r}
head(coords)
```
Using `network_neighbour` and the coordinate matrix we just produced, we can create a matrix representing links between nodes:

```{r}
nodes_degrees <- network_outdegree(nw, 1:N)
links_nodes <- rep(1:N, times=nodes_degrees)
links_indices <- unlist(mapply(seq, from=1, to=nodes_degrees))
links <- cbind(links_nodes, network_neighbour(nw, links_nodes, links_indices))
```

Each row of `links` represents a link and shows the coordinate of the two involved nodes:

```{r}
head(links)
```

The two matrices `coords` and `links` make it easy to plot the network with `ggplot2`. 

```{r}
ggplot() +
  geom_point(data=coords, aes(x=x, y=y),
             size=0.5) +
  geom_segment(data=cbind(a=as.data.frame(coords[links[, 1],]),
                          b=as.data.frame(coords[links[, 2],])),
               aes(x=a.x, y=a.y, xend=b.x, yend=b.y),
               linewidth=0.1) +
  theme_void()
```

### A `plot_network` function

Before we simulate the epidemic, we create a convenience function `plot_network` to visualize the state of the epidemic. The function uses `simulation_isinfected()` to query the infected state of each node and show non-infected and infected nodes in different colors. To make sure the plot encompasses the whole network, we use `network_bounds()` to query to bounds of our network in space and set the plotting limits 

```{r}
plot_network <- function(nw, sim=NULL) {
  # Node coordinates
  N <- network_size(nw)
  coords <- network_coordinates(nw, 1:N)
  colnames(coords) <- c('x', 'y')
  
  # Link coordinates
  nodes_degrees <- network_outdegree(nw, 1:N)
  links_nodes <- rep(1:N, times=nodes_degrees)
  links_indices <- unlist(mapply(seq, from=1, to=nodes_degrees))
  links <- cbind(links_nodes, network_neighbour(nw, links_nodes, links_indices))
  
  # Boundaries
  bounds <- network_bounds(nw)
  
  # Get data
  infected <- if (!is.null(sim))
    as.factor(simulation_isinfected(sim, 1:N))
  else
    as.factor(rep(FALSE, N))
  data_nodes <- cbind(as.data.frame(coords),
                      infected=infected)
  data_links <- cbind(a=as.data.frame(coords[links[, 1],]),
                      b=as.data.frame(coords[links[, 2],]))
  
  # Plot network
  ggplot() +
    geom_point(data=data_nodes, aes(x=x, y=y, color=infected, size=infected)) +
    geom_segment(data=data_links, aes(x=a.x, y=a.y, xend=b.x, yend=b.y),
                 linewidth=0.1) +
    scale_color_manual(breaks=c(FALSE,TRUE), values=c('cornflowerblue','orange')) +
    scale_size_manual(breaks=c(FALSE,TRUE), values=c(0.5, 2.0)) +
    labs(color="infected", size="infected") +
    lims(x=c(bounds[[1]][1], bounds[[2]][1]),
         y=c(bounds[[1]][2], bounds[[2]][2])) +
    theme_void() +
    theme(legend.position = 'none')
}
```

### Epidemics on spatial netwoks

We now create an animation depicting the spread of an epidemic on the lattice, starting from a single initial infection at the center of our lattice. During each iteration of the loop, the simulation is run until stopping time `t1` and the current epidemic state is plotted. To have the full list of events available later on, we combine the events reported by the calls to `simulation_run` and merge them into a large table at the end.

```{r, animation.hook='gifski', interval=0.1}
infection_time <- gamma_time(mean=6, var=4)
sim <- simulation(nw=nw, psi=infection_time)
simulation_addinfections(sim, nodes=L*floor(L/2) - L/2, times=c(0.0))

events <- list()
i <- 0
for(t1 in seq(from=0, to=250, by=5)) {
  i <- i + 1
  events[[i]] <- simulation_run(sim, stop=list(time=t1))
  print(plot_network(nw, sim))
}
events <- do.call(rbind, events)
```

Finally, we plot the number of infected nodes a function of time. In a log-plot, we see that the epidemic does not show a clear exponential regime

```{r warning=FALSE}
ggplot(events) +
  geom_line(aes(x=time, y=infected)) +
  scale_y_log10()
```

When we instead plot the trajectory in a log-log plot we clearly see the effect of the rather dense 2-dimensional embedding of the nodes -- instead of growing exponentially, the epidemic grows quadratically.

```{r warning=FALSE}
ggplot(events) +
  geom_line(aes(x=time, y=infected)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(slope=2, intercept=-1, linetype='dashed', color='blue')
```

## Temporal networks

Instead of the static network defined by `cubiclattice2d_network()`, we now consider a *temporal network*. On temporal networks, the set of links between nodes is a function of time, i.e. links appear and disappear, see `help(network_types)`. An example of a temporal network is `brownian_proximity_temporalnetwork()` where nodes diffuse randomly in two dimensions and are at any instant connected to all nodes within a certain radius. The density of points is chosen so that at any instant in time, each node has on average a prescribed number of neighbours `avg_degree`. Speed of movement is governed by two diffusion constants `D0` for susceptible nodes and `D1` for infected nodes which describe  the average squared distance per unit time that a node moves. 

```{r}
N2 <- 2500
nw2 <- brownian_proximity_temporalnetwork(
  N2, avg_degree=4, D0=1.0, D1=1.0, radius=1, dt=0.05)
```

The resulting network initially looks like this

```{r}
plot_network(nw2)
```

Since the network is random, node indices do not correspond to the position of nodes in the plane. However, we can use `network_coordinates` to find the node closest to the center of the plane.

```{r}
bounds2 <- network_bounds(nw2)
center2 <- 0.5*bounds2[[1]] + 0.5*bounds2[[2]]
coords2 <- network_coordinates(nw2, 1:N2)
d2 <- apply(coords2, MARGIN=1, function(x) sum((x - center2)**2))
CENTERNODE2 <- which.min(d2)
```

### Epidemics on temporal networks

Epidemics on temporal networks are the same way as on static networks, by creating a simulator object with `simulation()` and calling `simulation_run()`. However, `simulation_run()` will not not only change the epidemic state of nodes, but also evolve the network. Contrary to the situation on a static network, temporal networks are thus *modified* when `simulation_run()` is called. 

We now proceed as for the lattice and create an animation depicting the spread of an epidemic on a Brownian proximity network

```{r, animation.hook='gifski', interval=0.1}
infection_time2 <- gamma_time(mean=6, var=4)
sim2 <- simulation(nw=nw2, psi=infection_time2)
simulation_addinfections(sim2, nodes=CENTERNODE2, times=c(0.0))

events2 <- list()
i <- 0
for(t1 in seq(from=0, to=60, by=0.2)) {
  i <- i + 1
  events2[[i]] <- simulation_run(sim2, stop=list(time=t1, infected=N2))
  print(plot_network(nw2, sim2))
}
events2 <- do.call(rbind, events2)
```

### The effect of diffusion on epidemic growth

A log-plot of the number of infected node shows that contrary to the case of a lattice, the number of infected nodes initially grows exponentially again on the Brownian proximity network, despite the fact that the average degree of nodes is 4 for both networks.

```{r warning=FALSE}
ggplot(events2) +
  geom_line(aes(x=time, y=infected, color='Brownian proximity')) +
  geom_line(data=events, aes(x=time, y=infected, color='lattice')) +
  scale_y_log10() +
  lims(x=range(events2$time))
```


